---
title: "EDS 213 Final Project"
subtitle: "Data cleaning"
author: "Aakriti Poudel"
date: April 20, 2026
format:
  html:
    toc: true
    toc-depth: 3
    theme: cosmo
    code-fold: false
    embed-resources: true
---

```{r}
# forces R to always use the project root as the working directory
knitr::opts_knit$set(root.dir = here::here())
```


## Project background

This project uses the **What a Waste 3.0** database, published by the [World Bank Group](https://datacatalog.worldbank.org/search/dataset/0039597/what-a-waste-global-database). It is one of the most comprehensive global datasets on solid waste management, covering cities and countries across all income levels and world regions.

**Research Question:**

> *How does waste composition differ across income groups and regions? And does a country's
> waste composition predict its recycling or open dumping rates?*

---

## About the datasets

I am working with two separate datasets from the What a Waste 3.0 database.

### City dataset

- **File:** `what_a_waste_3_0_city_dataset.xlsx`
- **Coverage:** 262 cities across 163 countries
- **What it contains:** For each city, the dataset records the percentage breakdown of waste by material type (food, plastic, paper, glass, metal, etc.) and the percentage of waste sent to different treatment methods (recycling, open dumping, landfill, etc.)
- **Income groups:** Low Income (24 cities), Lower-Middle Income (79), Upper-Middle Income (81), High Income (78)
- **Regions:** Europe & Central Asia (89 cities), Sub-Saharan Africa (46), Latin America & the Caribbean (43), East Asia & Pacific (38), Middle East & North Africa (26), South Asia (14), North America (6)
- **Known data gaps:** Waste composition is missing for 62 of 262 cities. Recycling rates are missing for 148 cities, open dumping rates for 216 cities.

This reflects real-world reporting gaps — not errors in the data.

### Country dataset

- **File:** `what_a_waste_3_0_country_dataset.xlsx`
- **Coverage:** 217 countries (one row per country)
- **What it contains:** The same waste composition and treatment variables as the city dataset, but aggregated to the national level.
- **Income groups:** High Income (80 countries), Upper-Middle Income (57), Lower-Middle Income (50), Low Income (14). 16 countries have no income group assigned.
- **Known data gaps:** Composition data missing for ~38 countries. Recycling rates missing for 87 countries, open dumping rates for 165 countries.

### How the two datasets relate

Both datasets share the column `iso3c` — a standard three-letter country code (e.g., `"USA"`, `"KEN"`, `"BRA"`). This is the **join key** that links cities to their country. The relationship is **many-to-one**: one country row matches many city rows (for example, India has 5 cities in the city dataset).

The goal of this cleaning script is to produce two CSV files, one per dataset, where the `iso3c` join key is clean, consistent, and ready for SQL ingestion in Lab 3.

---

## Step 1: Load the libraries

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(readxl) # read Excel files directly into R
library(here)
library(janitor)

```

---

## Step 2: Load the raw data

Both files are Excel workbooks with multiple sheets. The data I need is in the sheets called `"City dataset"` and `"Country dataset"`.

```{r}
# read the city-level sheet
city_raw <- suppressMessages(
  read_excel(
  here::here("data", "raw", "what_a_waste_3_city_dataset.xlsx"),
  sheet = "City dataset") %>% 
  clean_names()
)

# read the country-level sheet
country_raw <- suppressMessages(
  read_excel(
  here::here("data", "raw", "what_a_waste_3_country_dataset.xlsx"),
  sheet = "Country dataset") %>% 
  clean_names()
)

# how many rows and columns are there
cat("City dataset:", nrow(city_raw), "rows,", ncol(city_raw), "columns\n")
cat("Country dataset:", nrow(country_raw), "rows,", ncol(country_raw), "columns\n")
```
---

## Step 3: Select only the columns we need

Both files have hundreds of columns. I keep only the columns relevant to my research question and rename them to short, clean names so they are easier to work with in future steps.

```{r}
# ── City dataset ──────────────────────────────────────────────────────────────

city_selected <- city_raw %>%
  select(
    # --- identifiers ---
    iso3c,        # 3-letter country code — JOIN KEY to country dataset
    country_name,
    city_name,
    city_code,    # unique city identifier — PRIMARY KEY for this table
    income_id,    # income group: LIC, LMIC, UMIC, HIC
    region_id,    # world region

    # --- waste composition (% of total waste by weight) ---
    food_pct    = composition_msw_food_organic_waste_percent,
    plastic_pct = composition_msw_plastic_percent,
    paper_pct   = composition_msw_paper_cardboard_percent,
    glass_pct   = composition_msw_glass_percent,
    metal_pct   = composition_msw_metal_percent,
    other_pct   = composition_msw_other_percent,

    # --- waste treatment outcomes (% of total waste by weight) ---
    recycling_pct = waste_treatment_recycling_percent,
    open_dump_pct = waste_treatment_open_dumpsite_percent
  )

# ── Country dataset ───────────────────────────────────────────────────────────

country_selected <- country_raw %>%
  select(
    # --- identifiers ---
    iso3c        = country_code,        # PRIMARY KEY for this table
    country_name = country_name,
    income_id    = income_group_waste_generation_year,
    region_id    = region,

    # --- waste composition (% of total waste by weight) ---
    food_pct    = composition_food_percent_weight_msw,
    plastic_pct = composition_plastic_percent_weight_msw,
    paper_pct   = composition_paper_cardboard_percent_weight_msw,
    glass_pct   = composition_glass_percent_weight_msw,
    metal_pct   = composition_metal_percent_weight_msw,
    other_pct   = composition_other_percent_weight_msw,

    # --- waste treatment outcomes (% of total waste by weight) ---
    open_dump_pct = treatment_open_dump_percent_weight_msw_generated,
    recycling_pct = treatment_recycling_percent_weight_msw_generated
  )

# confirm the column selection looks right
glimpse(city_selected)
glimpse(country_selected)
```

---

## Step 4: Remove the spurious header row (Country dataset)

When R reads the country Excel file, the very first data row contains column
name strings (like `"iso3c"`, `"income_id"`) instead of real values. This is
because the Excel file had a second header row embedded inside the data area.
That row is not real data and must be removed.

```{r}
# remove the row where iso3c literally contains the string "iso3c"
country_selected <- country_selected %>%
  filter(iso3c != "iso3c")

# verify it is gone, should print 0
cat("Leftover header rows (should be 0):",
    sum(country_selected$iso3c == "iso3c", na.rm = TRUE), "\n")
cat("Country rows after removal:", nrow(country_selected), "\n")
```

---

## Step 5: Fix the inconsistent income group label (Country dataset)

The `income_id` column in the country dataset contains one old label: `"UMC (2019)"`. This was used in a previous version of the dataset for upper-middle-income countries. I recode it to `"UMIC"` so it is consistent with the city dataset and with all other rows in the country dataset.

```{r}
# replace the old label with the current standard code
country_selected <- country_selected %>%
  mutate(income_id = if_else(income_id == "UMC (2019)", "UMIC", income_id))

# check what values remain, should only be LIC, LMIC, UMIC, HIC, and NA
country_selected %>%
  count(income_id, sort = TRUE)
```

---

## Step 6: Convert percentage oclumns to numeric

R read the composition and treatment columns as text (`character`) because the raw Excel file mixes numbers with occasional labels. I convert them all to numeric. Any value that cannot be converted will become `NA` automatically.

```{r}
# name all the percentage columns that need converting
numeric_cols <- c(
  "food_pct", "plastic_pct", "paper_pct",
  "glass_pct", "metal_pct",  "other_pct",
  "recycling_pct", "open_dump_pct"
)

# convert in both datasets using across()
city_selected <- city_selected %>%
  mutate(across(all_of(numeric_cols), as.numeric))

country_selected <- country_selected %>%
  mutate(across(all_of(numeric_cols), as.numeric))

# confirm all target columns are now numeric
cat("City — column types after conversion:\n")
city_selected %>% select(all_of(numeric_cols)) %>% sapply(class)

cat("\nCountry — column types after conversion:\n")
country_selected %>% select(all_of(numeric_cols)) %>% sapply(class)
```

---

## Step 7: Check for missing values

Missing values (`NA`) are common in this dataset because not every city or country has reported every variable. I count and display them here so I know what to expect in later analyses. I do **not** remove rows with missing values, the identifier columns on those rows are still valid and needed for joins.

```{r}
# missing values per column, city dataset
cat("=== Missing values: city dataset ===\n")
city_selected %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "column", values_to = "n_missing") %>%
  arrange(desc(n_missing)) %>%
  print(n = Inf)

# missing values per column, country dataset
cat("\n=== Missing values: country dataset ===\n")
country_selected %>%
  summarise(across(everything(), ~ sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "column", values_to = "n_missing") %>%
  arrange(desc(n_missing)) %>%
  print(n = Inf)
```

---

## Step 8: Check primary keys for duplicates

For a SQL join to work correctly, the primary key of each table must be unique. A duplicated primary key would cause rows to multiply during a join.

- `city_code` is the primary key of the city table
- `iso3c` is the primary key of the country table

```{r}
# check city_code, every city should appear exactly once
city_pk_dupes <- city_selected %>%
  count(city_code) %>%
  filter(n > 1)

cat("Duplicate city_code values (should be 0):", nrow(city_pk_dupes), "\n")
if (nrow(city_pk_dupes) > 0) print(city_pk_dupes)

# check iso3c in the country table, every country should appear exactly once
country_pk_dupes <- country_selected %>%
  count(iso3c) %>%
  filter(n > 1)

cat("Duplicate iso3c in country table (should be 0):", nrow(country_pk_dupes), "\n")
if (nrow(country_pk_dupes) > 0) print(country_pk_dupes)
```

---

## Step 9: Verify the join key overlaps across both tables

I check that every `iso3c` value in the city table has a matching row in the country table. If any city has no matching country, it would be dropped silently in an `INNER JOIN`.

```{r}
# unique iso3c values from each table
city_iso    <- unique(city_selected$iso3c)
country_iso <- unique(country_selected$iso3c)

# cities with and without a matching country row
matched   <- intersect(city_iso, country_iso)
unmatched <- setdiff(city_iso, country_iso)

cat("Unique countries represented in city table:", length(city_iso), "\n")
cat("Unique countries in country table:", length(country_iso), "\n")
cat("City-side iso3c WITH a country match:", length(matched), "\n")
cat("City-side iso3c WITHOUT a country match:", length(unmatched), "\n")

# print any unmatched values so we can investigate
if (length(unmatched) > 0) {
  cat("Unmatched iso3c values:", paste(unmatched, collapse = ", "), "\n")
}
```

All 163 city-side `iso3c` values have a matching country row, so no cities will be lost in a join. The 54 countries in the country table that have no cities are fine, they simply will not appear in a join result, which is expected.

---

## Step 10: Add readable income labels

The `income_id` codes are consistent but short. I add a full plain-language label column so plots and summary tables are easier to read.

Note: 16 countries in the country dataset have no income group assigned. Their `income_label` will be `NA`, which is the correct representation.

```{r}
# map short codes to readable labels
income_labels <- c(
  "LIC"  = "Low Income",
  "LMIC" = "Lower-Middle Income",
  "UMIC" = "Upper-Middle Income",
  "HIC"  = "High Income"
)

# add income_label to both datasets
city_selected <- city_selected %>%
  mutate(income_label = income_labels[income_id])

country_selected <- country_selected %>%
  mutate(income_label = income_labels[income_id])

# confirm the mapping
cat("City — income group counts:\n")
city_selected %>% count(income_id, income_label)

cat("\nCountry — income group counts (16 NAs expected):\n")
country_selected %>% count(income_id, income_label)
```

---

## Step 11: Save the cleaned datasets

Both datasets are clean and ready. I save them as CSV files so they can be loaded into a database in Lab 3.

```{r}
# preview each cleaned dataset before saving
cat("=== Cleaned city dataset — first 5 rows ===\n")
head(city_selected, 5)

cat("\n=== Cleaned country dataset — first 5 rows ===\n")
head(country_selected, 5)
```

```{r}
# write to CSV
write_csv(city_selected, "data/processed/city_cleaned.csv")
write_csv(country_selected, "data/processed/country_cleaned.csv")

cat("Saved: data/city_cleaned.csv —", nrow(city_selected), "rows,",
    ncol(city_selected), "columns\n")
cat("Saved: data/country_cleaned.csv —", nrow(country_selected), "rows,",
    ncol(country_selected), "columns\n")
```

---

## Cleaning summary

| Step | Problem found | What I did |
|------|--------------|------------|
| Select columns | 393 and 110 columns — most not needed | Kept 14–15 columns per table; renamed to short clean names |
| Remove header row | Row 1 in country dataset had column name strings as values | Filtered it out with `filter(iso3c != "iso3c")` |
| Fix income label | `"UMC (2019)"` appeared only in the country dataset | Recoded to `"UMIC"` with `if_else()` |
| Convert to numeric | Percentage columns came in as text from Excel | Used `as.numeric()` via `across()` |
| Missing values | Many NAs in composition and treatment columns | Counted and documented — not deleted |
| Primary key check | Duplicate PKs would break SQL joins | Confirmed 0 duplicates in both tables |
| Join key check | Unmatched `iso3c` would be dropped silently | Confirmed all 163 city-side values have a country match |
| Income labels | Short codes are hard to read in outputs | Added `income_label` column with full names |

The two cleaned CSV files, `city_cleaned.csv` and `country_cleaned.csv`, share the column `iso3c` as a join key and are ready for database ingestion.
